Generalized measurement invariance tests with application to factor analysis

### 2011-11-11 www/rev1/response.txt

Editor

Comment E.1. While I share the reviewers generally positive view of your paper, I find myself in the unusual position of being less enthusiastic about the present version of the paper than the reviewers seem to be.  One reason for this is that I think that your method is much less practical than the reviewers do, chiefly because in practical applications of invariance work, it is unusual to encounter a case in which V is continuous. The one exception to this situation is when V is age, but even then, in the context of employment discrimination for example, legal considerations often demand an arbitrary split on 40 years, or some other such threshold (which is admittedly arbitrary, but is demanded by the law).

Response E.1. Though we did not highlight it in the paper, the methods can be used with categorical V.  We have also developed a new strategy for handling ordinal V and briefly describe it in the Discussion.  Additionally, we would argue that applications of invariance do not use continuous V because methods for handling continuous V are still emerging (e.g., the moderated factor models discussed by Reviewer 1).  We believe that our proposed methods will encourage invariance research involving continuous V.

Comment E.2. The second point is related to issues raised by Reviewers one and four.  In any invariance analyses, some elements of theta are allowed to vary across groups freely because they are not part of the invariance constraints dictated by equation 7.  In the factor model, the factor means and factor covariance elements are not constrained.  How is this fact represented in your method?  While I saw your treatment of partial invariance (i.e., hold some elements to invariance while letting other elements go free), you seem in that context to be testing invariance in all of the free elements.  Does your method allow one to permit some elements to be unconstrained, while others are
invariant, without looking for violations of invariance in the free elements?  If it does, we need to see how this works.

Response E.2. This comment focuses on the fact that our methods generally work from a "reduced model" standpoint, whereby all model parameters are constrained to be equal and we examine the stability of parameter estimates across individuals (or across groups).  When groups are not known in advance (e.g., when V is continuous), this appears to be the only way to proceed.

However, if groups are known in advance, it is possible to employ the proposed methods in tandem with a multiple-group model.  The multiple-group model can then handle both unconstrained and invariant parameters, because scores can be computed for all model parameters.  We now illustrate this approach in the data example. DONE

Comment E.3. You also did not mention how model identification constraints, which in the invariance case are sometimes invariance constraints, are handled.

Response E.3. The proposed tests are based on scores associated with all free parameters.  Thus, it is straightforward to handle parameters that are constrained to be equal to one another.  However, this comment may also refer to the fact that one's choice of identification constraint will influence tests of other free parameters.  We now briefly describe this issue in the "Use in Practice" section of the Discussion.  DONE

Comment E.4. I think that we need to have a real example, as suggested by Reviewer one.  This would allow you to answer my second point above, and also show that your method can give answers that are sensible in context.  With a real example, you could drop the artificial one (but keep the simulations). Having a real example would also help to illustrate that there are real applications in which your method would prove useful.

Response E.4. We have dropped the artificial example and added a real example. DONE

Reviewer #1

Comment 1.1. The section on the asymptotic properties of the Beta-function on pages 11-14 was still quite hard to follow. The authors may want to expand upon some of the more technical aspects of these models to allow more readers to grasp how they would do the job in their given application.

Response 1.1. We have revised the theory development and added extra detail to the text.  If some specific concepts are still unclear, we are happy to revisit this section.  DONE

Comment 1.2. There are clear links between the current approach based on an observed continuous grouping variable V and the factor mixture approach. In factor mixture modeling, the grouping variable that characterizes the subgroups that differ in terms of particular model parameters are also typically unknown (i.e., latent classes), but class membership can be predicted on the basis of observed predictor(s). So the model in Figure 1 and the scenario of 2 subgroups that the authors describe on pages 17-20 is essentially a factor mixture model with two latent classes in which age predicts class membership. Therefore, the authors should take steps to relate their modeling approach to factor mixture modeling.

Response 1.2. We agree and have now described our methods' relationship to mixture models in the Discussion.  Some advantages of our approach include the facts that it does not require a priori specification of the number of latent classes or a priori specification of the way by which age impacts class membership.  DONE

Comment 1.3. Similarly, there exist a family of moderation-type factor models that may serve the same purpose of studying failures of invariance in common factor models with respect to (unknown) "grouping" variables. For instance, Dylan Molenaar, Conor Dolan and colleagues have recently worked out CFA models that were based on Shaun Purcell's work in which certain parameters are moderated by either observed or unobserved variables. These models too could be applied to the important problems related to invariance with respect to unknown groups although they may not work as well as do mixture models and the current approach when indeed there is some step function that describes the violations of invariance. The existence of these alternative approaches by no means diminishes the potential value of the current developments, but it does mean that the authors should discuss the relation between their approach and other existing methods.

Response 1.3. We now describe factor analysis models employing moderation in the "Comparison to other methods" section of the discussion.  DONE

Comment 1.4. The authors chose to simulate data only under models that did not include the mean structure. Since violations of measurement invariance with real data are most often related to the mean structure (to be sure due to larger power associated with modeling first-order moments), it may be more realistic and interesting to have scenarios in which such types of failures of invariance (called uniform measurement bias) occur.

Response 1.4. Mean parameters were estimated and tested in the simulation models, though they were unstructured.  Mean structures are often made possible by constraining parameters across groups, and our simulation focuses on a case where groups cannot be predefined.  Thus, we have not modified the model in the simulation.  In the real data example that we have added, however, the mean structure is utilized. DONE

Comment 1.5. Also, many "nuisance" variable that are hypothesized to affect indicators, say, in cognitive testing (e.g., see Wicherts et al.'s work on measurement invariance and stereotype threat) in a given multi-group CFA are expected to linearly affect these indicators. Now I can come up with a host of scenarios in which the measurement bias is approximately linear and so can be modeled in simple terms because the regression on V results in an additional factor in the model.  Surely I don't expect the authors to deal with all of such potential scenarios' of failures of measurement invariance in CFA. However, I do wish to point out that underlying the current approach is a specific kind of measurement bias that is expected to be discontinuous while many failures of invariance are expected to be continuous and approximately linear. Therefore, the authors should go into more detail on the nature of the measurement bias that the approach is trying to uncover. Their approach may be generic, but it is not really general in the sense that many failures of strict factorial invariance can be modeled in more straightforward and traditional ways. The step function in Eq. 10 should be discussed in more detail and may be supplemented by a hypothetical example of why it occurs in linear factor models.

Response 1.5. The step function from Equation 10 is employed as an illustrative example because it highlights the connection between nested model comparisons and the proposed techniques. Also, the maxLM statistic from Equation 17 has particularly high power for this scenario. However, the proposed tests typically have non-trivial power for all (non-constant) deviation patterns theta(v_i) (considered in Equation 9).  We have now provided more details on these issues in the Discussion section (and tied the details to Comment/Response 1.3).  DONE

Comment 1.6. On page 3, the authors state that failures of measurement invariance render scales useless, but this statement may be overly strong. Specifically, scales that are only partially invariant  may still be useful. In addition, the statement that "the scales are the sole cause of the differences" appears to suggest that failures of invariance can explain all group differences in observed test scores, which is seldom the case. Failures of measurement invariance do not preclude the existence of group differences in the distribution of latent factors scores.

Response 1.6. We agree that these statements were too strong and have revised accordingly.  DONE

Comment 1.7. When presenting the definition of invariance on page 5, the authors may wish to indicate that the definition (due originally to Mellenbergh, 1989) implies (as shown by Meredith, 1993) that not all of the parameters in the CFA are invariant but only the measurement parameters. So factor means, factor covariances, and (in structural models) factor regression weights are not expected to be invariant under this definition of invariance.
Another issue that could be discussed is that non-normality in the data may lead to spurious findings in the current approach just like they do in regular multi-group confirmatory analyses and in factor mixture modeling (e.g., see work by Gitta Lubke and Michael Neale).

Response 1.7. We now describe the invariance definitions in more detail, and we also clarify that our methods are not necessarily robust to violations of normality.  DONE

Comment 1.8. Finally, it may be interesting to apply the method to real data and to compare the outcomes from more traditional modeling approaches. After all, in real life the exact nature of the measurement bias with respect to V is not know. One interesting approach may be to use data from experiments like in the stereotype threat framework that was used by Wicherts et al. (who I am sure will be happy to share raw data for this purpose).

Response 1.8. See Response E.4.  DONE

Reviewer #3

Comment 3.1. The result from Hjort and Koning is a nice trick to rewrite the asymptotic representation of the likelihood estimator as an empirical process. This then relates the cumulative sums of the likelihood estimator to a Brownian bridge. The weak convergence of the monitoring process is of course asymptotic. The properties of the tests in the current manuscript are investigated decently enough for medium to large sample sizes (n = 100 to 500). However, for small samples no results are obtained. Given the relatively small model in the simulations, I would expect a relatively small sample size of n = 50 to be included in the simulations. The main reason is that this may show a point where the power functions show a "breakdown."

Response 3.1. We now include the n=50 condition in the simulations, which indeed exhibits a breakdown in the sense that power is not monotonic with respect to violation magnitude.  DONE

Comment 3.2. The weak convergence of the monitoring process is proved in the paper by Hjort and Koning in two cases: (i) for theta_0 the true parameter, and (ii)   for local alternatives O(1/sqrt(n)) from theta_0. In the latter case the monitoring process is shown to be a Brownian bridge including a constant (bias) even when using -E{u'u}, Fisher's information (which is consistent up to O(1/n)). I would imagine this is important since model misspecification (measurement invariance) is the topic of the manuscript. This result allows the authors to to determine exactly what the result of misspecification is. It fits well with their line of reasoning, which I agree with. The misspecification is determined by a deviation h/sqrt(n) of the true parameter. This corresponds exactly to the misspecification in the simulations. Measurement invariance can be imagined as a latent variable acting on the factor loadings, and can therefore be incorporated in the parameter theta. If there were no
measurement invariance, theta_v woold be 0, but in the case of misspecification, it is exactly the same as that of Hjort and Koning. I would recommend that the authors use this result by Hjort and Koning in the manuscript to show analytically what happens to the distribution and tests.

Response 3.2. We have now incorporated these results in a new section, titled "Local alternatives".  We also discuss these results in the context of the simulation, showing the impact of our imposed measurement invariance violation on the Brownian bridge.  DONE

We addressed all minor issues identified by Reviewer 3.  We have a response to one of these issues:

Comment 3.3. p. 7 second paragraph: Assuming that the model is correctly specified is bonus when the second point above is incorporated. I think this would improve the manuscript.

Response 3.3. Violations of distributional assumptions may still cause problems with the proposed tests, even if the "structure" of the model (e.g., parameter constraints, included variables, etc) is correct.  We have modified this section of text accordingly.  DONE

Reviewer #4

Comment 4.1. Page 6/7:	It could be emphasized that a known cutpoint ν means “observed heterogeneity” and an unknown cutpoint ν means “unobserved heterogeneity”.

Response 4.1. We have now added this detail.  DONE

Comment 4.2. Page 9:	The authors should give a reference or a short explanation of the “tied-down Bessel process”. Readers might be unfamiliar with the process.

Response 4.2. We now note that Andrews (1993) provides results and references relevant to Bessel processes.  DONE

Comment 4.3. Page 10:	To test for changes “along” V means that V is always continuous (also mentioned in the first paragraph). What happens if V is an ordered categorical variable? Would that be a special case? Adding a reference to the discussion section would be helpful.

Response 4.3. We have added some discussion on the use of categorical V.  Also see Response E.1.

Comment 4.4. Page 19:	The overall k=19 model parameter consists of six factor loadings, six measurement errors and one covariance between the factors. The authors refer to the six means of the manifest variables. Usually, in factor models mean differences of the factors are of research interest. The authors should give some information why they considered the manifest means instead of the latent mean differences.

Response 4.4. Factor means can be estimated if one is constraining other model parameters to be equal across groups.  In our new data example, we do estimate factor means and use them in the proposed tests.  In our simulation (and in the example that was removed), we use a scenario where no grouping variables exist (i.e., there is only a continuous V).  In this scenario, it seems difficult to estimate factor means prior to model estimation because we cannot constrain parameters across groups.  We have added some extra detail to the simulation section on our use of a single-group model.  DONE

Comment 4.5.	The simulations later on focus on invariance of factor loadings. Basicly, the covariance structure in the simulations are analyzed with the factor model of Figure 1. So, means are not included in the analyses (as far as I read the simulation results correctly).

Response 4.5. Mean parameters are included in the casewise likelihood function, though they were unmodeled in the simulation and thus omitted from Figure 1.  The casewise likelihood function is needed to make the proposed tests possible.  Also see Comment/Response 1.5.  DONE

Comment 4.6. Page 25: 	Implications of the study of measurement invariance in factor analysis to general structural equation models could be added with comments to mean differences (see comment for page 19).

Response 4.6. We have added some brief discussion surrounding extensions to structural equation models.  DONE

Comment 4.7. Table 1:	Table captions should be moved to the bottom of the table. (Abbreviations: …)

Response 4.7. We have moved the Abbreviations part of the captions.
DONE

### 2011-11-11 www/paper.Rnw
Addressed more reviewer comments, including:   - Added commentary on the n=50 simulation condition   - Addressed the issue of identification conditions impacting     the test results
Also added a draft set of responses to reviewers in the rev1 directory.


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@74 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

\begin{figure}
\caption{Three test statistics of~\eqref{eq:exh0} (set of
  group-specific parameters)
  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
  using data from Study 1 of Wicherts et al.\ (2005).
  Solid, horizontal lines represent critical values
  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\setkeys{Gin}{width=\textwidth}
<<gefp-4-1, fig=TRUE, height=8, width=6>>=
## add small amount of random noise to duplicated GPAs, so that everything
## works with zoo.  (Without this, one can still obtain DM and CvM test 
##                   results but not supLM results.)
set.seed(1090)
ties <- duplicated(d$gpa)
d$gpa[ties] <- runif(sum(ties), d$gpa[ties]-.01, d$gpa[ties]+.01)
## empirical fluctuation processes (4 vs 1 par)
gefp_4_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
gefp_1_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)

@@

\subsection{Method}
Data were generated using the model in Figure~\ref{fig:famod}, with
$V$ being designed to reflect student ages.
For the measurement invariance violation, ``young'' students (13--15)
had smaller factor loadings 
than ``old'' students (16--18).
Sample size and magnitude
of measurement 
invariance violation were manipulated to examine power:
we examined power to detect 
invariance violations across three sample sizes ($n=100,
200, 500$) and 17~magnitudes of violations.  These violations
involved the younger students' values of $\{\lambda_{11},
\lambda_{21}, \lambda_{31}\}$ deviating from the older students'
values by $d$ 
times the parameters' asymptotic standard errors (scaled by $\sqrt{n}$),
with $d = 0, 0.25, 0.5, \dots, 4$.  The 0-standard error condition was
used to study Type I error rate.

@@

Figure~\ref{fig:simres} displays power curves as a function of violation
magnitude, with panels for each combination of sample size ($n$) $\times$
number of parameters being tested ($k^{\ast}$).  
Separate curves are drawn for the double-max test (solid lines), 
the Cram{\'e}r-von Mises test (dashed lines), and the $\max \mathit{LM}$ (dotted lines).
One can generally observe that simultaneous tests of all 19~parameters
result in decreased power, with the tests performing more similarly at
the larger sample sizes.  The tests distinguish themselves from one
another when only the three factor loadings are tested, with the
Cram\'{e}r-von Mises test having the most power, followed by $\max \mathit{LM}$,
followed by the double-max test.  This advantage decreases with
increases in sample size.  
Table~\ref{tab:simres} presents the same results as
Figure~\ref{fig:simres}, but it is easier to see exact power
magnitudes in the table.
The table shows that the
power advantage of the Cram\'{e}r-von
Mises test can be as large as 0.1, most notably when three parameters
are being tested.
  It also shows that the Cram\'{e}r-von Mises test generally
has true Type-I error rates, with the double-max test being somewhat
conservative and the $\max \mathit{LM}$ test being slightly liberal.

@@

\begin{figure}
\caption{Simulated power curves for the double-max test (solid), 
  Cram\'{e}r-von Mises test (dashed), and $\max \mathit{LM}$ test (dotted)
  across three sample sizes $n$, two subsets of tested parameters $k^*$, and
  measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$). 
  See Table~\ref{tab:simres} for the underlying numeric values (using a
  subset of nine violation magnitudes).}
\label{fig:simres}
\setkeys{Gin}{width=\textwidth}
<<mzsim-xyplot, fig=TRUE, height=7.5, width=7>>=
## visualization
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(c("CvM", "max LM", "DM"), points = FALSE, lines = TRUE)
mykey$lines$lty <- c(2, 3, 1)
mykey$lines$lwd <- c(1.2, 2, 1.2)
print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub,
  type = "l", lwd = c(1, 1, 2), ylim = c(0, 1), key = mykey,
  xlab = "Violation Magnitude", ylab = "Power"))
@
\end{figure}

\begin{table*}
\caption{Simulated power for three test statistics
  across three sample sizes $n$, nine magnitudes of measurement
  invariance violations, and two subsets of tested
  parameters $k^*$. 
  See Figure~\ref{fig:simres} for a visualization (using all 17~violation magnitudes).}
\label{tab:simres}
\begin{center}
\begin{tabular}{lrlrrrrrrrrr}
  \hline
   & & & \multicolumn{9}{l}{Violation Magnitude (SE)} \\
<<mz_sim-table, results=tex>>=
## labeling
levels(mz_sim_sub$pars) <- gsub("k* = ", "", levels(mz_sim_sub$pars), fixed = TRUE)
levels(mz_sim_sub$nobs) <- gsub( "n = ", "", levels(mz_sim_sub$nobs), fixed = TRUE)
levels(mz_sim_sub$test) <- c("$\\mathit{DM}$", "$\\mathit{CvM}$", "$\\max \\mathit{LM}$")

@@

@@

@@

More generally, all the approaches described above can be distinguished
from the proposed tests in that they require estimation of a new model
with increased complexity (due to the moderator/covariate).
The proposed tests, on the other hand, are of the
``posthoc'' variety, relying only on results calculated during the
original model estimation.  While no method clearly dominates 
across all situations, use of the proposed tests can at least reduce 
some of the technical issues associated with estimating and
interpreting models of greater complexity.  This may not be a
good reason to use the tests, but it is a consideration that is often
meaningful in practice.

@@

The proposed tests also 
readily extend to other popular psychometric models.  
Instead of studying measurement invariance in factor
analysis, the tests can be used to generally 
study the stability of structural equation model parameters across
observations.  For 
example, it would be possible to assess whether paths between
latent variables are stronger for some individuals than for
others.  Secondly, 
the tests may be extended to study differential item functioning (DIF)
in item response models (e.g., \citeNP{KopZei10}, who focused on recursive
partitioning of Rasch models).  
Traditional DIF methods are similar to those for factor analysis in
that subgroups must be specified in advance.
%% The tests proposed here can detect subgroups automatically, however, offering a unified way of
%% studying both  measurement invariance in factor analysis and differential item
%% functioning in item response.
While factor-analytic measurement invariance methods and DIF methods
have developed 
largely independently of one another, 
some treatments \cite{Mcd99} and recent research \cite{StaChe06} have
sought to unify the methods.  Extensions of the proposed tests can
support this endeavor.

\subsection{Summary}
We have outlined a family of 
stochastic process-based parameter instability tests from theoretical
statistics and applied them to the issue of measurement invariance in psychometrics.
The paper included both theoretical development and 
study of the tests' performance.  The tests were found to have good
properties via simulation, making them useful for many psychometric
applications.  More generally, the tests help 
solve standing problems in
measurement invariance research and provide many avenues for
future research, both through extensions of the tests within a
factor-analytic context and through application of the tests to new
models.

>>>>

# Refit model with new identification constraints
wdh0.1 <- wdhfit2(d)

@@

\begin{figure}
\caption{Three test statistics of~\eqref{eq:exh0} (set of
  group-specific parameters)
  and~\eqref{eq:exh0.2} (intercept only) across values of student GPA, 
  using data from Study 1 of Wicherts et al.\ (2005).
  Solid, horizontal lines represent critical values
  at $\alpha = 0.05$, and the dotted, horizontal lines (second row)
  represent values of the Cram\'{e}r-von Mises test statistic.}
\label{fig:cusumex}
\setkeys{Gin}{width=\textwidth}
<<gefp-4-1, fig=TRUE, height=8, width=6>>=
## add a small amount of random noise to duplicated GPAs, so that everything
## works with gefp().  (Without this, one can still obtain DM and CvM test 
##                      results but not supLM results.)
set.seed(1090)
ties <- duplicated(d$gpa)
d$gpa[ties] <- runif(sum(ties), d$gpa[ties]-.01, d$gpa[ties]+.01)
## empirical fluctuation processes (4 vs 1 par)
gefp_4_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
gefp_1_info <- gefp(wdh0, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)

# processes for second identification constraints
gefp_4_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 15:18, vcov = info.wdhfit, sandwich = FALSE)
gefp_1_info_1 <- gefp(wdh0.1, fit = NULL, order.by = d$gpa[d$grouping=="exp:minority"], parm = 18, vcov = info.wdhfit, sandwich = FALSE)

@@

\subsection{Method}
Data were generated using the model in Figure~\ref{fig:famod}, with
$V$ being designed to reflect student ages.
For the measurement invariance violation, ``young'' students (13--15)
had smaller factor loadings 
than ``old'' students (16--18).
Sample size and magnitude
of measurement 
invariance violation were manipulated to examine power:
we examined power to detect 
invariance violations across four sample sizes ($n=50, 100,
200, 500$) and 17~magnitudes of violations.  These violations
involved the younger students' values of $\{\lambda_{11},
\lambda_{21}, \lambda_{31}\}$ deviating from the older students'
values by $d$ 
times the parameters' asymptotic standard errors (scaled by $\sqrt{n}$),
with $d = 0, 0.25, 0.5, \dots, 4$.  The 0-standard error condition was
used to study Type I error rate.

@@

Figure~\ref{fig:simres} displays power curves as a function of violation
magnitude, with panels for each combination of sample size ($n$) $\times$
number of parameters being tested ($k^{\ast}$).  
Separate curves are drawn for the double-max test (solid lines), 
the Cram{\'e}r-von Mises test (dashed lines), and the $\max \mathit{LM}$ (dotted lines).
One can generally observe that simultaneous tests of all 19~parameters
result in decreased power, with the tests performing more similarly at
the larger sample sizes.  The tests distinguish themselves from one
another when only the three factor loadings are tested, with the
Cram\'{e}r-von Mises test having the most power, followed by $\max \mathit{LM}$,
followed by the double-max test.  
This advantage decreases at larger sample sizes.  
At $n=50$, generally regarded as a
small sample size for factor analysis, the power functions break down
and are not monotonic with respect to violation magnitude.

Table~\ref{tab:simres} presents a subset of the results displayed in 
Figure~\ref{fig:simres}, but it is easier to see exact power
magnitudes in the table.
The table shows that the
power advantage of the Cram\'{e}r-von
Mises test can be as large as 0.1, most notably when three parameters
are being tested.
  It also shows that the Cram\'{e}r-von Mises test generally
has true Type-I error rates, with the double-max test being somewhat
conservative and the $\max \mathit{LM}$ test being slightly liberal.

@@

\begin{figure}
\caption{Simulated power curves for the double-max test (solid), 
  Cram\'{e}r-von Mises test (dashed), and $\max \mathit{LM}$ test (dotted)
  across four sample sizes $n$, two subsets of tested parameters $k^*$, and
  measurement invariance violations of 0--4 standard errors (scaled by $\sqrt{n}$). 
  See Table~\ref{tab:simres} for the underlying numeric values (using a
  subset of nine violation magnitudes).}
\label{fig:simres}
\setkeys{Gin}{width=\textwidth}
<<mzsim-xyplot, fig=TRUE, height=7.5, width=7>>=
## visualization
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(c("CvM", "max LM", "DM"), points = FALSE, lines = TRUE)
mykey$lines$lty <- c(2, 3, 1)
mykey$lines$lwd <- c(1.2, 2, 1.2)
print(xyplot(power ~ diff | pars + nobs, group = ~ test, data = mz_sim_sub,
  type = "l", lwd = c(1, 1, 2), ylim = c(0, 1), key = mykey,
  xlab = "Violation Magnitude", ylab = "Power"))
@
\end{figure}

\begin{table*}
\caption{Simulated power for three test statistics
  across four sample sizes $n$, nine magnitudes of measurement
  invariance violations, and two subsets of tested
  parameters $k^*$. 
  See Figure~\ref{fig:simres} for a visualization (using all 17~violation magnitudes).}
\label{tab:simres}
\begin{center}
\begin{tabular}{lrlrrrrrrrrr}
  \hline
   & & & \multicolumn{9}{l}{Violation Magnitude (SE)} \\
<<mz_sim-table, results=tex>>=
## labeling
levels(mz_sim_sub$pars) <- gsub("k* = ", "", levels(mz_sim_sub$pars), fixed = TRUE)
levels(mz_sim_sub$nobs) <- gsub( "n = ", "", levels(mz_sim_sub$nobs), fixed = TRUE)
levels(mz_sim_sub$test) <- c("$\\mathit{DM}$", "$\\mathit{CvM}$", "$\\max \\mathit{LM}$")

@@

<<<constraint_study>>=
tests <- c("maxBB", "meanL2BB", "supLM(0.1)")

p.diffs <- matrix(NA, 2*length(tests), 2)

for (i in 1:length(tests)){
  p.diffs[i,1] <- unlist(sctest(gefp_4_info, functional=
                                eval(parse(text=tests[i])))[2])
  p.diffs[i,2] <- unlist(sctest(gefp_4_info_1, functional=
                                     eval(parse(text=tests[i])))[2])
  }
for (i in (length(tests)+1):(2*length(tests))){
  p.diffs[i,1] <- unlist(sctest(gefp_1_info, functional=
                                     eval(parse(text=tests[i-length(tests)])))[2])
  p.diffs[i,2] <- unlist(sctest(gefp_1_info_1, functional=
                                   eval(parse(text=tests[i-length(tests)])))[2])
}
@

@@

Another difference between the proposed tests and traditional tests
involves the fact that, for the proposed tests, one's choice of
identification constraints can influence tests of the remaining
parameters.  This essentially reflects the fact that, while the model
fits equally well regardless of identification constraint, the
free model parameters adjust themselves to the specific identification
constraint chosen.  To examine this issue in more detail, we refit the model
used in the Example section, using an alternative identification
constraint (i.e., fixing a different factor loading at 1).  Across the 6
tests presented in the example section (see, e.g.,
Figure~\ref{fig:cusumex}), the largest difference in p-values for the
two identification constraints 
was~$\Sexpr{round(max(abs(p.diffs[,2] - p.diffs[,1])),3)}$.  This is not a large difference, though further
study is warranted.

@@

More generally, all the approaches described above can be distinguished
from the proposed tests in that they require estimation of a new model
of increased complexity (due to the moderator/covariate).
The proposed tests, on the other hand, are of the
``posthoc'' variety, relying only on results calculated during the
original model estimation.  While no method clearly dominates 
across all situations, use of the proposed tests can at least reduce 
some of the technical issues associated with estimating and
interpreting models of greater complexity.  This alone is not a
good reason to use the tests, but it is a consideration that is often
meaningful in practice.

@@

The proposed tests also 
readily extend to other popular psychometric models.  
For example, the tests can be used to generally 
study the stability of structural equation model parameters across
observations.  Under this framework, it is possible to assess whether
paths between 
latent variables are stronger for some individuals than for
others.  Further, in applying the proposed tests to factor mean
parameters (as was done in the example), implications of measurement
invariance violations for structural equation modeling may be better
understood.

The proposed tests also naturally extend to the study of differential
item functioning (DIF) 
in item response models (e.g., \citeNP{KopZei10}, who focused on Rasch
models).   
Traditional DIF methods are similar to those for factor analysis in
that subgroups must be specified in advance.
%% The tests proposed here can detect subgroups automatically, however, offering a unified way of
%% studying both  measurement invariance in factor analysis and differential item
%% functioning in item response.
While factor-analytic measurement invariance methods and DIF methods
have developed 
largely independently of one another, the methods can certainly be
treated from a unified perspective (e.g.,
\citeNP{Mcd99,Mil11,StaChe06}).   
The tests proposed here were designed with this perspective in mind.

\subsection{Summary}
We have outlined a family of 
stochastic process-based parameter instability tests from theoretical
statistics and applied them to the issue of measurement invariance in psychometrics.
The paper included theoretical development, an applied example, and 
study of the tests' performance.  The tests were found to have good
properties via simulation, making them useful for many psychometric
applications.  More generally, the tests help 
solve standing problems in
measurement invariance research and provide many avenues for
future research.  This can happen both through extensions of the tests
within a factor-analytic context and through application of the tests to new
models.
